---
title: "Multivariate Analysis in SAS"
author: "Aftab A Sorwar"
format: 
  revealjs:
    theme: serif
    width: 1200
    height: 800
    css: styles.css 
    include-after-body: 
      - text: |
          <style>
          .reveal h1.title {
            font-size: 2em; 
            white-space: nowrap;
          }
          </style>
    smaller: False
    preview-links: true
    toc: true
    toc-depth: 2
    code-fold: show
    transition: slide #fade
    scrollable: true
    slide-number: true
    chalkboard: true
    code-line-numbers: true
execute:
  echo: true
  warning: false
  message: false
---

## **Today’s goal**

By the end of this workshop you can:

-   Explain what **multivariate data** is!
-   Choose a method based on the Objective.
-   Run each method in **R** and interpret results

## **What is multivariate data?**

In a **Multivariate case** we observe multiple variables(responses) per unit.

**Example:**

-   a patient's health records tracking cholesterol, blood pressure, and weight
-   car data (MPG, horsepower, weight)
-   house features (size, bedrooms, price)
-   dataset assessing crop yield, Soil pH, Nitrogen Level, Irrigation Amount and Avg. Temperature

## **Univariate vs multivariate (quick picture)**

-   **Univariate**: one response (e.g., compare yield for different Soil pH levels)
-   **Multivariate**: multiple responses\
    (e.g., compare yield across different Soil pH levels, Nitrogen Levels, Irrigation Amounts, and Avg.) Temperature)

Multivariate methods help avoid: - running many separate tests - missing joint patterns

## **Why multivariate methods?**

::: notes
To see the big picture. The more variable the better it is.
:::

-   Many variables are **correlated**

-   We want **summary patterns**, **groups**, or **group differences** across several outcomes

## **Workflow checklist (always)**

1.  Identify the **Objective**
2.  Choose variables (numeric vs categorical)
3.  Clean data: missing values, outliers
4.  Decide whether to **standardize** (usually yes for PCA & clustering)
5.  Fit model + interpret output
6.  Communicate the results in simple words

## **Choosing technique based on objective**

<details name="my-lists">

<summary>**Predict or Explain**</summary>

-   Multiple Regression

-   Logistic Regression

-   Discriminant Analysis

</details>

<details name="my-lists">

<summary>**Reduce complexity (Dimension reduction)**</summary>

-   [Principal Component Analysis (PCA)](#PCA){style="color: red;"}

-   Factor Analysis (FA)

</details>

<details name="my-lists">

<summary>**Find Groups (unsupervised learning)**</summary>

-   [Cluster Analysis (CA)](#CA){style="color: red;"}

-   Multidimensional Scaling

</details>

<details name="my-lists">

<summary>**Understand relationships between sets of variables**</summary>

-   Canonical Correlation Analysis (CCA)

-   Structural Equation Modeling

</details>

<details name="my-lists">

<summary>**Compare groups on multiple outcomes**</summary>

-   Multivariate Analysis of Variance (MANOVA)

</details>

## **PCA** {#PCA}

::: notes
"In the next 2 minutes, I'm going to show you how PCA can take complex agricultural data and make it simple enough to understand at a glance. Let's dive in!" TIMING: 10 seconds
:::

**A statistical technique that transforms correlated variables into a smaller set of uncorrelated variables called principal components.**

::: notes
Imagine a pencil • Side view: Long & detailed • End view: Just a circle

PCA finds the "side view" The angle that shows the most variation and information
:::

### Why use PCA 

:::: {.columns style="font-size: .8em;"}

::: {.column width="50%"}
* Dimensionality Reduction

* Remove Redundancy

* Feature Engineering
:::

::: {.column width="50%"}
* Data Visualization

* Noise Reduction

* Speed up Analysis
:::

::::

### When to use PCA

:::: {.columns style="font-size: 0.7em;"}

::: {.column width="50%"}
[**Good Scenarios:**]{style="color: Green;"}

1. Many correlated variables (r > 0.7)
2. Need to visualize complex data
3. All variables are numeric/continuous
4. Variables measured on similar scales
5. Exploratory data analysis
6. Pre-processing for ML models

:::

::: {.column width="50%"}
[**Not Recommended When:**]{style="color: red;"}

1. Variables are independent (not correlated)
2. Need to interpret original variables
3. Categorical/binary variables only
4. Small sample size (n < 50)
5. Missing data is widespread
6. Variables on very different scales (unless standardized)

:::

::::


:::{aside}
**Tip:** Check correlation matrix first! If most correlations < 0.3, PCA may not help much.
:::


## **CA** {#CA}

**Cluster Analysis is an unsupervised method that groups observations** (fields, farms, samples, people) so that:

:::: {.columns style="font-size: 0.75em;"}
* items within the same cluster are similar, and

* items in different clusters are different

based on the variables chosen (usually numeric variables).

::::

### Why use CA 

:::: {.columns style="font-size: 0.8em;"}

* discover types/profiles in your data (e.g. “field types”)

* simplify a complex dataset into a few understandable groups

* support decision-making (e.g., “these fields behave similarly, manage them similarly”)

* identify unusual observations (outliers often show up as small or isolated clusters)

::::



### When to use CA
:::: {.columns style="font-size: 0.7em;"}
* do not have labeled groups already (no “true classes”)

* have several variables and suspect there are hidden patterns

* goal is exploration: “What kinds of observations exist here?”

::::

## **Example**

We’ll use a dataset:

::: notes
We have data from 180 agricultural fields across a few Nebraska counties. For each field we recorded soil test results (N, P, K, pH, organic matter, clay %, soil moisture), a vegetation index (NDVI) from imagery during the growing season, and the end-of-season yield. We also have field characteristics like crop type, irrigation type, and management (conventional vs cover crop). The goal is to use multivariate methods to summarize soil + vegetation conditions and find field types.
:::

-   We have data from 180 agricultural fields across a few Nebraska counties.

-   For each field, we recorded soil test results (N, P, K, pH, organic matter, clay %, soil moisture), a vegetation index (NDVI) from imagery during the growing season, and the end-of-season yield. We also have field characteristics like crop type, irrigation type, and management (conventional vs cover crop).

-   The goal is to use multivariate methods to summarize soil + vegetation conditions and find field types.

## **Loading data in R**

```{r}
url <- "https://raw.githubusercontent.com/aftabsorwar/STAT-930-workshop/refs/heads/main/multivariate_workshop_dataset.csv"
dat <- read.csv(url)

# Save `dat` into your current working directory
write.csv(dat, file = "multivariate_workshop_dataset.csv", row.names = FALSE)

# (Optional) check where it was saved
getwd()

# make group variables factors
dat$Crop <- factor(dat$Crop)
dat$Management <- factor(dat$Management)

# Quick look
str(dat)
head(dat)
```

:::: {.columns style="font-size: .8em;"}
Short description of the variable:

-   **FieldID:** Unique ID for each field.

-   **Crop:** Crop type grown *(Corn / Soybean / Wheat)*.

-   **Management:** Practice type *(Conventional vs CoverCrop)*.

-   **SoilN:** Soil nitrogen level (mg/kg). *Higher often means more available nitrogen for plants.*

-   **SoilP:** Soil phosphorus level (ppm). *Supports root development and early growth.*

-   **OrgMatter:** Organic matter (%). *A soil health indicator. Improves fertility and water-holding capacity.*

-   **Moisture:** Soil moisture (%). *Amount of water in the soil.*

-   **NDVI:** A satellite/drone-based vegetation index (0–1) that measures how green the crop is. *Larger values generally mean more plant biomass and stronger vegetation cover.*
::::


## **PCA in R**

```{r}
# Create a vector of the numeric column names we want for PCA
num_vars <- c("SoilN", "SoilP", "OrgMatter", "Moisture", "NDVI")

# Create a numeric-only data frame X containing just those variables
X <- dat[, num_vars]

# Confirm that X contains only numeric columns
str(X)

# Count missing values in each column
colSums(is.na(X))
```


```{r}
# Calculate correlation matrix
cor_matrix <- cor(X)
print(round(cor_matrix, 2))

# Visualize correlations
# install.packages("corrplot")
library(corrplot)

corrplot(cor_matrix, 
         method = "color", 
         type = "upper",
         addCoef.col = "black", 
         number.cex = 0.7,
         tl.col = "black", 
         tl.srt = 45,
         title = "Correlation Matrix: Field Variables",
         mar = c(0, 0, 2, 0))
```

```{r}
# Fit PCA using prcomp:
# center=TRUE subtracts the mean
# scale.=TRUE divides by standard deviation (standardizes variables)
pca_fit <- prcomp(X, center = TRUE, scale. = TRUE)

# Print a basic summary of the PCA result
summary(pca_fit)
```

:::: {.columns style="font-size: .7em;"}

PC1 and PC2 are new summary variables created by PCA. 

They are made from your original variables, but they are designed to:

1. Summarize the data with fewer dimensions

2. Capture the biggest patterns in the data

    a.PC1 captures the largest overall variation (the main trend).

    b.PC2 captures the second-largest variation, separate from PC1.

3. these 2 are uncorrelated (i.e. PC2 is built so it does not repeat what PC1 already explains.)
::::

```{r}
# pca_fit$sdev contains the standard deviations of each principal component
# Squaring gives variance of each PC
pc_var <- pca_fit$sdev^2

# Convert to percentage of variance explained (PVE)
pve <- pc_var / sum(pc_var)
pve_pct <- pve*100
pve_pct <- round(pve_pct, 3)
pve_pct
cumulative_pve_pct <- cumsum(pve_pct)
cumulative_pve_pct
```

:::: {.columns style="font-size: .8em;"}
PC1 explains 55%, PC2 explains 22%. Together about 78% of the variation

::::


```{r}
# Plot proportion variance explained for each PC
plot(pve, type = "b",
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained")
```

```{r}
# Show loadings for the first two PCs
round(pca_fit$rotation[, 1:2], 3)
```


:::: {.columns style="font-size: .7em;"}
A loading is a correlation-like weight, so its value is between −1 and +1

What the size means

* Close to +1 or −1 → strong contribution of that variable to the PC

* Close to 0 → little contribution

* The sign does not matter, magnitude matters.

::::

:::: {.columns style="font-size: .7em;"}
For our example:

* PC1 mainly represents a “soil fertility / soil health”.
    - Fields differ on PC1 mostly because of SoilN, SoilP, and organic matter.

* PC2 mainly represents a “water + crop greenness”.
    - Fields differ on PC2 mostly because of soil moisture and NDVI.
::::

```{r}
biplot(pca_fit,
       cex = 0.6,        # point/label size
       scale = 0,        # keeps interpretation simple
       xlabs = rep(".", nrow(X)))  # hides observation labels (too crowded)

```

::: notes
Blue dots = individual fields (each point is one field)

Red arrows = your original variables (SoilN, SoilP, OrgMatter, Moisture, NDVI)

PC1 (x-axis) and PC2 (y-axis) are the two main PCA directions

:::

:::: {.columns style="font-size: .7em;"}

This biplot shows that PC1 mainly reflects soil fertility (SoilN, SoilP, OrgMatter), while PC2 mainly reflects moisture and crop greenness (Moisture, NDVI). Fields spread along these two gradients.


* A field located toward the left tends to have higher SoilN/SoilP/OrgMatter

* A field located higher up tends to have higher Moisture and NDVI

* Fields near the center are closer to average on all variables
::::

## **CA in R**

```{r}
# Standardize to mean 0 and sd 1
X_scaled <- scale(X)


# Elbow method: try k from 1 to 8
set.seed(930)
wss <- numeric(8)

for (k in 1:8) {
  km <- kmeans(X_scaled, centers = k, nstart = 25)
  wss[k] <- km$tot.withinss
}
```

```{r}
plot(1:8, wss, type = "b",
     xlab = "Number of clusters (k)",
     ylab = "Total within-cluster sum of squares")

```

```{r}
set.seed(930)
km3 <- kmeans(X_scaled, centers = 3, nstart = 25)

# Add cluster labels to the dataset
dat$Cluster <- factor(km3$cluster)

# How many fields per cluster?
table(dat$Cluster)

```

```{r}
# Cluster means (interpretation table)
cluster_means <- aggregate(X, by = list(Cluster = dat$Cluster), mean)
cluster_means

```

```{r}
# PCA only for plotting (not required for clustering)
pca <- prcomp(X_scaled, center = FALSE, scale. = FALSE)
scores <- as.data.frame(pca$x[, 1:2])

plot(scores$PC1, scores$PC2,
     pch = 19,
     xlab = "PC1", ylab = "PC2",
     main = "PCA space (without cluster)")

plot(scores$PC1, scores$PC2,
     col = dat$Cluster, pch = 19,
     xlab = "PC1", ylab = "PC2",
     main = "Clusters shown in PCA space")
legend("topright", legend = levels(dat$Cluster),
       col = 1:length(levels(dat$Cluster)), pch = 19)

```

::: notes
Points with the same color are grouped as similar field profiles. Separation means clusters differ.
:::


:::: {.columns style="font-size: .7em;"}
Clustering helps us find field types without a response variable. Then we summarize each cluster using the mean soil and NDVI values to create an interpretable story.
::::

